Assignment: Course Reproducible-Research : Project 1
========================================================
  
### DATA SOURCE  
Data is downloaded from the source (https://www.coursera.org/learn/reproducible-research/peer/gYyPt/course-project-1) on 7/JAN/2015 and unzipped for processing.
#### Data Description  
Data is from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

### Loading and preprocessing the data

Data is loaded using `read.csv` function with the following code.

```{r, echo=TRUE}
data <- read.csv("D:/r_learn/data_scientist/reproducible/week1/repdata-data-activity/activity.csv", colClasses = c("integer", "Date", "factor"))
data$month <- as.numeric(format(data$date, "%m"))
datanoNA <- na.omit(data)
rownames(datanoNA) <- 1:nrow(datanoNA)
head(datanoNA)

```



### What is mean total number of steps taken per day?  

#### Make a histogram of the total number of steps taken each day 
```{r,echo=TRUE}  
library(ggplot2)
ggplot(datanoNA, aes(date, steps)) + geom_bar(stat = "identity", colour = "red", fill = "red", width = 0.7) + facet_grid(. ~ month, scales = "free") + labs(title = "Histogram of Total Number of Steps Taken Each Day", x = "Date", y = "Total number of steps")
```


#### Mean total number of steps taken per day:  

```{r, echo=TRUE}
totalSteps <- aggregate(datanoNA$steps, list(Date = datanoNA$date), FUN = "sum")$x
mean1= mean(totalSteps)
mean1

```


#### Median total number of steps taken per day:

```{r,echo=TRUE}

median1=median(totalSteps)
median1
```
 
### What is the average daily activity pattern?  

```{r,echo=TRUE}
avgSteps <- aggregate(datanoNA$steps, list(interval = as.numeric(as.character(datanoNA$interval))), FUN = "mean")
names(avgSteps)[2] <- "meanOfSteps"

ggplot(avgSteps, aes(interval, meanOfSteps)) + geom_line(color = "steelblue", size = 0.8) + labs(title = "Time Series Plot of the 5-minute Interval", x = "5-minute intervals", y = "Average Number of Steps Taken")
```


#### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r,echo=TRUE}

avgSteps[avgSteps$meanOfSteps == max(avgSteps$meanOfSteps), ]
```
 
#### The total number of rows with NAs:

```{r,echo=TRUE}

sum(is.na(data))
```


#### Filling in all of the missing values in the dataset.  

My strategy is to use the mean for that 5-minute interval to fill each NA value in the steps column.
New dataset that is equal to the original dataset but with the missing data filled in.


```{r,echo=TRUE}

newData <- data 
for (i in 1:nrow(newData)) {
    if (is.na(newData$steps[i])) {
        newData$steps[i] <- avgSteps[which(newData$interval[i] == avgSteps$interval), ]$meanOfSteps
    }
}

newData$steps<-ceiling(newData$steps)

head(newData)
```

#### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.  

```{r,echo=TRUE}

ggplot(newData, aes(date, steps)) + geom_bar(stat = "identity",
                                             colour = "red",
                                             fill = "red",
                                             width = 0.7) + facet_grid(. ~ month, scales = "free") + labs(title = "Histogram of Total Number of Steps Taken Each Day (no missing data)", x = "Date", y = "Total number of steps")
```
 
#### Mean total number of steps taken per day in modified DATAset  

```{r,echo=TRUE}

newTotalSteps <- aggregate(newData$steps, 
                           list(Date = newData$date), 
                           FUN = "sum")$x
mean2 <- mean(newTotalSteps)
mean2
```

#### Median total number of steps taken per day in modified DATAset 

```{r,echo=TRUE}

median2 <- median(newTotalSteps)
```

#### Compare them with the two before imputing missing data: 

```{r,echo=TRUE}

mean1 - mean2

median1 - median2
```


So, after imputing the missing data both mean and median is increased.

### Are there differences in activity patterns between weekdays and weekends?

EXCERSIE: Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.


```{r,echo=TRUE}
newData$weekdays <- factor(format(newData$date, "%A"))
levels(newData$weekdays)

```

Creating Levels

```{r,echo=TRUE}
levels(newData$weekdays) <- list(weekday = c("Monday", "Tuesday",
                                             "Wednesday", 
                                             "Thursday", "Friday"),
                                 weekend = c("Saturday", "Sunday"))
levels(newData$weekdays)


```

Calculating number of each level

```{r,echo=TRUE}
table(newData$weekdays)
```

EXCERCISE:
Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).


```{r,echo=TRUE}

avgSteps <- aggregate(newData$steps, 
                      list(interval = as.numeric(as.character(newData$interval)), 
                           weekdays = newData$weekdays),
                      FUN = "mean")
names(avgSteps)[3] <- "meanOfSteps"
library(lattice)
xyplot(avgSteps$meanOfSteps ~ avgSteps$interval | avgSteps$weekdays, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Number of steps")
```

